\chapter{Quotient Groups and Homomorphisms}

\section{Definitions and Examples}

Let $G$ and $H$ be groups.

\Exercise1 Let $\varphi\colon G\to H$ be a homomorphism and let $E$ be
a subgroup of $H$. Prove that $\varphi^{-1}(E)\leq G$ (i.e., the
preimage or pullback of a subgroup under a homomorphism is a
subgroup). If $E\trianglelefteq H$ prove that
$\varphi^{-1}(E)\trianglelefteq G$. Deduce that
$\ker\varphi\trianglelefteq G$.
\begin{proof}
  Note that $\varphi(1) = 1\in E$ so $\varphi^{-1}(E)$ is
  nonempty. Suppose $a, b \in \varphi^{-1}(E)$, so that
  $\varphi(a) = x$ and $\varphi(b) = y$ for some $x,y\in E$. Then,
  since $\varphi$ is a homomorphism, we have
  \begin{equation*}
    \varphi(ab^{-1}) = \varphi(a)\varphi(b)^{-1} = xy^{-1} \in E,
  \end{equation*}
  which shows that $ab^{-1}\in\varphi^{-1}(E)$. By the subgroup
  criterion, this shows that $\varphi^{-1}(E)\leq G$.

  Now suppose that $E$ is a normal subgroup of $H$. Let $g\in G$ and
  $n\in\varphi^{-1}(E)$. Then $\varphi(g) = h$ for some $h\in H$ and
  $\varphi(n) = x$ for some $x\in E$. We have
  \begin{align*}
    \varphi(gng^{-1})
    &= \varphi(g)\varphi(n)\varphi(g)^{-1} \\
    &= hxh^{-1}.
  \end{align*}
  But $hxh^{-1}\in E$ since $E\trianglelefteq H$, so
  $gng^{-1}\in\varphi^{-1}(E)$. The choice of $g$ and $n$ were
  arbitrary, so this shows that $\varphi^{-1}(E)\trianglelefteq G$.

  Lastly, if we let $E$ be the trivial subgroup of $H$, then the above
  shows that $\ker\varphi = \varphi^{-1}(E) \trianglelefteq G$ since
  the trivial subgroup is always normal.
\end{proof}

\Exercise2 Let $\varphi\colon G\to H$ be a homomorphism of groups with
kernel $K$ and let $a,b\in\varphi(G)$. Let $X\in G/K$ be the fiber
above $a$ and let $Y$ be the fiber above $b$, i.e.,
$X = \varphi^{-1}(a)$, $Y = \varphi^{-1}(b)$. Fix an element $u$ of
$X$ (so $\varphi(u) = a$). Prove that if $XY = Z$ in the quotient
group $G/K$ and $w$ is any member of $Z$, then there is some $v\in Y$
such that $uv = w$.
\begin{proof}
  Let $v = u^{-1}w$. We want to show that $v\in Y$, or
  $\varphi(v) = b$. Since $\varphi$ is a homomorphism and since
  $Z = \varphi^{-1}(ab)$, we have
  \begin{align*}
    \varphi(v) &= \varphi(u^{-1}w) \\
               &= \varphi(u)^{-1}\varphi(w) \\
               &= a^{-1}(ab) \\
               &= (a^{-1}a)b \\
               &= b.
  \end{align*}
  So $v\in Y$ as required.
\end{proof}

\Exercise3 Let $A$ be an abelian group and let $B$ be a subgroup of
$A$. Prove that $A/B$ is abelian. Give an example of a non-abelian
group $G$ containing a proper normal subgroup $N$ such that $G/N$ is
abelian.
\begin{solution}
  Let $a_1B, a_2B \in A/B$, where $a_1,a_2\in A$. Since $A$ is
  abelian, we have
  \begin{equation*}
    a_1Ba_2B = (a_1a_2)B = (a_2a_1)B = a_2Ba_1B.
  \end{equation*}
  Therefore $A/B$ is abelian.

  For the second part of the problem, let $G$ be the non-abelian
  dihedral group $D_8$ and let $N$ the proper normal subgroup
  $\gen{r^2}$. In the text it was shown that $G/N \cong V_4$, the
  Klein four-group, which is abelian. Therefore $G/N$ is abelian even
  though $G$ is not.
\end{solution}

\Exercise4
\label{exercise:quotient-group:coset-exponent}
Prove that in the quotient group $G/N$,
$(gN)^\alpha = g^\alpha N$ for all $\alpha\in\Z$.
\begin{proof}
  First, if $\alpha = 0$, then $(gN)^0 = 1N = g^0N$, so the statement
  is true in this case. We also know by Proposition~5 that
  $(gN)^{-1} = g^{-1}N$. So it will suffice to prove the statement
  only for positive $\alpha$, which we will do by induction on
  $\alpha$.

  The case where $\alpha = 1$ is trivial. For the inductive step,
  suppose that the statement $(gN)^k = g^kN$ holds for some particular
  $k\geq1$. Then
  \begin{align*}
    (gN)^{k+1}
    &= (gN)^kgN \\
    &= g^kNgN \\
    &= (g^kg)N \\
    &= g^{k+1}N.
  \end{align*}
  By induction, we conclude that $(gN)^\alpha = g^\alpha N$ for all
  $\alpha\geq1$, so the proof is complete.
\end{proof}

\Exercise5 Use the preceding exercise to prove that the order of the
element $gN$ in $G/N$ is $n$, where $n$ is the smallest positive
integer such that $g^n\in N$ (and $gN$ has infinite order if no such
positive integer exists). Give an example to show that the order of
$gN$ in $G/N$ may be strictly smaller than the order of $g$ in $G$.
\begin{solution}
  Fix an element $gN$ in $G/N$. First, if possible, let $n$ be the
  smallest positive integer such that $g^n\in N$. Then $g^nN =
  1N$. So, by the previous exercise, we know that $(gN)^n = 1N$. This
  shows that $\ord{gN}\leq n$. On the other hand, if $m$ is any
  positive integer with $(gN)^m = 1N$ then, using the previous
  exercise again, $g^mN = 1N$ so that $g^m\in N$. Since $n$ is the
  smallest positive integer with $g^n\in N$, this shows that
  $\ord{gN}\geq n$, which completes the proof for the case of finite
  order.

  Next, suppose that there is no such $n$. Then for each positive
  integer $k$, $g^k\not\in N$. If $gN$ were to have finite order, say
  $(gN)^m = 1N$, then the previous exercise would show that
  $g^m\in N$, giving a contradiction. This shows that $gN$ has
  infinite order, which completes the proof.

  Lastly, for the example, consider $G = Z_4$, the cyclic group of
  order $4$. Let $x$ be a generator for $G$ and take
  $N = \gen{x^2} = \{1, x^2\}$. Now the element $x^2$ has order $2$ in
  $G$, but $x^2N = 1N$ has order $1$ in $G/N$.
\end{solution}

\Exercise6 Define $\varphi\colon\R^\times\to\{\pm1\}$ by letting
$\varphi(x)$ be $x$ divided by the absolute value of $x$. Describe the
fibers of $\varphi$ and prove that $\varphi$ is a homomorphism.
\begin{solution}
  The fiber above $1$ is the positive reals, and the fiber above $-1$
  is the negative reals.

  Let $x,y\in\R^\times$ be arbitrary. Then
  \begin{equation*}
    \varphi(xy)
    = \frac{xy}{\abs{xy}}
    = \frac{x}{\abs{x}}\cdot\frac{y}{\abs{y}}
    = \varphi(x)\varphi(y),
  \end{equation*}
  so $\varphi$ is a homomorphism.
\end{solution}

\Exercise7 Define $\pi\colon\R^2\to\R$ by $\pi((x,y)) = x + y$. Prove
that $\pi$ is a surjective homomorphism and describe the kernel and
fibers of $\pi$ geometrically.
\begin{solution}
  For any $(x_1,y_1), (x_2,y_2)\in\R^2$, we have
  \begin{align*}
    \pi((x_1,y_1) + (x_2,y_2))
    &= \pi((x_1 + x_2, y_1 + y_2)) \\
    &= (x_1 + x_2) + (y_1 + y_2) \\
    &= (x_1 + y_1) + (x_2 + y_2) \\
    &= \pi((x_1,y_1)) + \pi((x_2,y_2)),
  \end{align*}
  so $\pi$ is a homomorphism. And for any $x\in\R$, we have
  $\pi((x, 0)) = x + 0 = x$, so $\pi$ is also surjective.

  $\ker\pi$ is simply the diagonal line whose equation is $x + y =
  0$. And for $a\in\R$, the fiber over $a$ is the line with equation
  $x + y = a$, which is just a translate of the kernel.
\end{solution}

\Exercise8 Let $\varphi\colon\R^\times\to\R^\times$ be the map sending
$x$ to the absolute value of $x$. Prove that $\varphi$ is a
homomorphism and find the image of $\varphi$. Describe the kernel and
the fibers of $\varphi$.
\begin{solution}
  For any $x,y\in\R^\times$ we have
  \begin{equation*}
    \varphi(xy) = \abs{xy} = \abs{x}\abs{y} = \varphi(x)\varphi(y)
  \end{equation*}
  and $\varphi$ is a homomorphism. Its image is $\R^+$, the positive
  reals.

  The kernel of $\varphi$ is the set $\{-1,1\}$, since
  $\varphi(\pm1) = 1$ and no other real number has an absolute value
  of $1$. Likewise, the fiber over the real number $a$ is $\{-a, a\}$.
\end{solution}

\Exercise9 Define $\varphi\colon\C^\times\to\R^\times$ by
$\varphi(a + bi) = a^2 + b^2$. Prove that $\varphi$ is a homomorphism
and find the image of $\varphi$. Describe the kernel and the fibers of
$\varphi$ geometrically (as subsets of the plane).
\begin{solution}
  Let $a + bi$ and $c + di$ be any members of $\C^\times$. Then
  \begin{align*}
    \varphi((a + bi)(c + di))
    &= \varphi((ac - bd) + (ad + bc)i) \\
    &= (ac - bd)^2 + (ad + bc)^2 \\
    &= a^2c^2 - 2abcd + b^2d^2 + a^2d^2 + 2abcd + b^2c^2 \\
    &= a^2(c^2 + d^2) + b^2(c^2 + d^2) \\
    &= (a^2 + b^2)(c^2 + d^2) \\
    &= \varphi(a + bi)\varphi(c + di),
  \end{align*}
  and $\varphi$ is a homomorphism. Let $a+bi\in\C^\times$. Since $a$
  and $b$ cannot both be zero, $a^2 + b^2 > 0$. So
  $\im\varphi\subseteq\R^+$. But for any $a\in\R^+$, we have
  $\varphi(\sqrt{a} + 0i) = a$ and we see that $\im\varphi = \R^+$.

  The kernel of $\varphi$ is the set
  \begin{equation*}
    \ker\varphi = \{a + bi\in\C^\times \mid a^2 + b^2 = 1\}.
  \end{equation*}
  This is a circle of radius 1 centered at the origin in the complex
  plane. For $a\in\R^+$, the fiber over $a$ is the circle of radius
  $\sqrt{a}$.
\end{solution}

\Exercise{10} Let $\varphi\colon\Z/8\Z\to\Z/4\Z$ by
$\varphi(\bar a) = \bar a$. Show that this is a well defined,
surjective homomorphism and describe its fibers and kernel explicitly
(showing that $\varphi$ is well defined involves the fact that
$\bar a$ has a different meaning in the domain and range of
$\varphi$).
\begin{solution}
  To show that $\varphi$ is well defined we need to show that any
  choice of representative for a particular congruence class in
  $\Z/8\Z$ will produce the same congruence class in $\Z/4\Z$ under
  $\varphi$. Suppose then that $\bar a = \bar b$, where
  $\bar a,\bar b\in\Z/8\Z$. Then for some integer $k$, $a = b + 8k$
  and we have
  \begin{align*}
    \varphi(\bar a)
    &= \varphi(\overline{b + 8k}) \\
    &= \overline{b + 8k} \\
    &= \overline{b + 4(2k)} \\
    &= \bar b \\
    &= \varphi(\bar b)
  \end{align*}
  and $\varphi$ is well defined. It is also a homomorphism since
  \begin{align*}
    \varphi(\bar a + \bar b)
    &= \varphi(\overline{a + b}) \\
    &= \overline{a + b} \\
    &= \bar a + \bar b \\
    &= \varphi(\bar a) + \varphi(\bar b).
  \end{align*}
  And it is clearly surjective.

  The fibers of $\varphi$ are
  \begin{align*}
    \varphi^{-1}(\bar0) &= \{\bar0, \bar4\}, \\
    \varphi^{-1}(\bar1) &= \{\bar1, \bar5\}, \\
    \varphi^{-1}(\bar2) &= \{\bar2, \bar6\}, \\
    \varphi^{-1}(\bar3) &= \{\bar3, \bar7\},
  \end{align*}
  and $\ker\varphi = \varphi^{-1}(\bar0) = \{\bar0, \bar4\}$.
\end{solution}

\Exercise{11} Let $F$ be a field and let
\begin{equation*}
  G = \left\{
    \begin{pmatrix}
      a & b \\
      0 & c
    \end{pmatrix}
    \;\;\middle|\;\;
    a,b,c\in F,\;\; ac\neq0
  \right\}
  \leq GL_2(F).
\end{equation*}
\begin{enumerate}
\item Prove that the map
  \begin{equation*}
    \varphi\colon
    \begin{pmatrix}
      a & b \\ 0 & c
    \end{pmatrix}
    \mapsto a
  \end{equation*}
  is a surjective homomorphism from $G$ onto $F^\times$. Describe the
  fibers and kernel of $\varphi$.
  \begin{solution}
    First, for any $a\in F^\times$, we have
    \begin{equation*}
      \varphi\left(
      \begin{pmatrix}
        a & 1 \\ 0 & 1
      \end{pmatrix}
      \right) = a,
    \end{equation*}
    so $\varphi$ is surjective. And for any $a,b,c,d,e,f\in F$ with
    $ac\neq0$ and $df\neq0$, we have
    \begin{align*}
      \varphi\left(\begin{pmatrix} a & b \\ 0 & c \end{pmatrix}
              \begin{pmatrix} d & e \\ 0 & f \end{pmatrix}\right)
      &= \varphi\left(\begin{pmatrix} ad & ae + bf \\
          0 & cf \end{pmatrix}\right) \\
      &= ad \\
      &= \varphi\left(\begin{pmatrix} a & b \\ 0 & c \end{pmatrix}\right)
         \varphi\left(\begin{pmatrix} d & e \\ 0 & f \end{pmatrix}\right),
    \end{align*}
    so $\varphi$ is a homomorphism.

    For each $a\in F^\times$, the fiber over $a$ is given by
    \begin{equation*}
      \varphi^{-1}(a) =
      \left\{
        \begin{pmatrix}
          a & x \\
          0 & y
        \end{pmatrix}
        \;\;\middle|\;\;
        x,y\in F,\;\;y\neq0
      \right\},
    \end{equation*}
    with the kernel being the fiber above $1$.
  \end{solution}

\item Prove that the map
  \begin{equation*}
    \psi\colon
    \begin{pmatrix}
      a & b \\ 0 & c
    \end{pmatrix}
    \mapsto (a,c)
  \end{equation*}
  is a surjective homomorphism from $G$ onto
  $F^\times\times F^\times$. Describe the fibers and kernel of $\psi$.
  \begin{solution}
    The proof is entirely similar to the proof for $\varphi$ in the
    previous part of the problem and is omitted here. The fiber over
    $(a,c)$ is
    \begin{equation*}
      \psi^{-1}((a,c)) =
      \left\{
        \begin{pmatrix}
          a & x \\
          0 & c
        \end{pmatrix}
        \;\;\middle|\;\;
        x\in F
      \right\},
    \end{equation*}
    with $\ker\psi = \psi^{-1}((1,1))$.
  \end{solution}

\item Let
  \begin{equation*}
    H = \left\{
      \begin{pmatrix}
        1 & b \\ 0 & 1
      \end{pmatrix}
      \;\middle|\;
      b\in F
    \right\}.
  \end{equation*}
  Prove that $H$ is isomorphic to the additive group $F$.
  \begin{proof}
    Define $\rho\colon H\to F$ by
    \begin{equation*}
      \rho\left(
        \begin{pmatrix}
          1 & b \\
          0 & 1
        \end{pmatrix}
      \right) = b.
    \end{equation*}
    Then $\rho$ is a bijection since it has the obvious two-sided
    inverse $\rho^{-1}\colon F\to H$ given by
    \begin{equation*}
      \rho^{-1}(b) =
      \begin{pmatrix}
        1 & b \\
        0 & 1
      \end{pmatrix},
    \end{equation*}
    and it is a homomorphism since
    \begin{align*}
      \rho\left(
      \begin{pmatrix}
        1 & b_1 \\ 0 & 1
      \end{pmatrix}
      \begin{pmatrix}
        1 & b_2 \\ 0 & 1
      \end{pmatrix}
      \right)
      &= \rho\left(
        \begin{pmatrix}
          1 & b_1 + b_2 \\ 0 & 1
        \end{pmatrix}
      \right) \\
      &= b_1 + b_2 \\
      &= \rho\left(
      \begin{pmatrix}
        1 & b_1 \\ 0 & 1
      \end{pmatrix}\right) +
      \rho\left(\begin{pmatrix}
        1 & b_2 \\ 0 & 1
      \end{pmatrix}
      \right).
    \end{align*}
    Therefore $\rho$ is an isomorphism and $H\cong F$.
  \end{proof}
\end{enumerate}

\Exercise{12}
\label{exercise:quotient-group:r-to-s1-kernel}
Let $G$ be the additive group of real numbers, let $H$
be the multiplicative group of complex numbers of absolute value $1$
(the unit circle $S^1$ in the complex plane) and let
$\varphi\colon G\to H$ be the homomorphism
$\varphi\colon r\mapsto e^{2\pi ir}$. Draw the points on a real line
which lie in the kernel of $\varphi$. Describe similarly the elements
in the fibers of $\varphi$ above the points $-1$, $i$, and
$e^{4\pi i/3}$ of $H$.
\begin{solution}
  The kernel of $\varphi$ is simply $\Z$, since $e^{2\pi ir} = 1$ if
  and only if $r$ is an integer:
  \begin{center}
    \begin{tikzpicture}
      \draw [->] (-5.6,0) -- (5.6,0);
      \foreach \i in {-5,...,5}{
        \draw [fill] (\i,0) circle [radius=1.5pt];
        \node [below] at (\i,0) {$\i$};
      }
    \end{tikzpicture}
  \end{center}

  The specified fibers are
  \begin{align*}
    \varphi^{-1}(-1)
    &= \frac12 + \Z = \left\{k + \frac12 \;\middle|\; k\in\Z\right\}, \\
    \varphi^{-1}(i)
    &= \frac14 + \Z = \left\{k + \frac14 \;\middle|\; k\in\Z\right\}, \\
    \varphi^{-1}(e^{4\pi i/3})
    &= \frac23 + \Z = \left\{k + \frac23 \;\middle|\; k\in\Z\right\}.
      \qedhere
  \end{align*}
\end{solution}

\Exercise{13} Repeat the preceding exercise with the map $\varphi$
replaced by the map $\varphi\colon r\mapsto e^{4\pi ir}$.
\begin{solution}
  In this case the kernel is $\frac12\Z$:
  \begin{center}
    \begin{tikzpicture}
      \draw [->] (-5.6,0) -- (5.6,0);
      \foreach \i in {-5,...,5}{
        \draw [fill] (\i,0) circle [radius=1.5pt];
        \node [below] at (\i,0) {$\i$};
      }
      \foreach \i in {-5,...,4}{
        \draw [fill] (\i + 1/2,0) circle [radius=1.5pt];
      }
    \end{tikzpicture}
  \end{center}

  The fibers are
  \begin{align*}
    \varphi^{-1}(-1)
    &= \frac14 + \frac12\Z
      = \left\{\frac{2k+1}4 \;\middle|\; k\in\Z\right\}, \\
    \varphi^{-1}(i)
    &= \frac18 + \frac12\Z
      = \left\{\frac{4k+1}8 \;\middle|\; k\in\Z\right\}, \\
    \varphi^{-1}(e^{4\pi i/3})
    &= \frac13 + \frac12\Z
      = \left\{\frac{3k+2}6 \;\middle|\; k\in\Z\right\}.
      \qedhere
  \end{align*}
\end{solution}

\Exercise{14} Consider the additive quotient group $\Q/\Z$.
\begin{enumerate}
\item Show that every coset of $\Z$ in $\Q$ contains exactly one
  representative $q\in\Q$ in the range $0\leq q<1$.
  \begin{proof}
    Let $t\in\Q$ be arbitrary. Write $t = m/n$ in lowest terms. We may
    use the division algorithm to find unique integers $q$ and $r$
    with $0\leq r<n$ such that $m = nq + r$. Then $t = q + r/n$, where
    $0\leq r/n<1$. Then we have
    \begin{equation*}
      t + \Z = \{t + k\mid k\in\Z\}
      = \left\{\frac{r}n + (k+q) \;\middle|\; k\in\Z\right\}
      = \frac{r}n + \Z.
    \end{equation*}
    From the uniqueness of $q$ and $r$, it follows that $r/n$ is the
    only representative of $t + \Z$ that is in the range $[0,1)$.
  \end{proof}
\item Show that every element of $\Q/\Z$ has finite order but that
  there are elements of arbitrarily large order.
  \begin{proof}
    Again, let $t\in\Q$ be arbitrary, with $t = m/n$ for integers $m$
    and $n$, with $n>0$. Then
    \begin{equation*}
      n(t + \Z) = nt + \Z = m + \Z = 0 + \Z,
    \end{equation*}
    so $t + \Z$ has finite order (note that the first equality follows
    from Exercise~\ref{exercise:quotient-group:coset-exponent}).

    Given any positive integer $k$, the coset $1/k + \Z$ has order
    $k$. Since $k$ can be made arbitrarily large, we see that $\Q/\Z$
    contains elements of arbitrarily large order.
  \end{proof}
\item Show that $\Q/\Z$ is the torsion subgroup of $\R/\Z$.
  \begin{proof}
    We need to show that the only elements in $\R/\Z$ having finite
    order belong to $\Q/\Z$. So suppose $r$ is an irrational
    representative of a coset having finite order $n$. Then
    \begin{equation*}
      nr + \Z = n(r + \Z) = 0 + \Z.
    \end{equation*}
    This implies that $nr\in\Z$, say $m = nr$. Then $r = m/n$ and
    $r\in\Q$, contradicting our choice of $r$. Therefore $r$ cannot
    have finite order. So $\Q/\Z$ is indeed the torsion subgroup of
    $\R/\Z$.
  \end{proof}
\item Prove that $\Q/\Z$ is isomorphic to the multiplicative group of
  roots of unity in $\C^\times$.
  \begin{proof}
    Let $S^1$ be the unit circle in the complex plane and let $G$ be
    the multiplicative group of the roots of unity. From
    Exercise~\ref{exercise:quotient-group:r-to-s1-kernel} we know that
    $\R/\Z \cong S^1$, as we can exhibit the explicit isomorphism
    $\varphi$ by
    \begin{equation*}
      \varphi(r + \Z) = e^{2\pi ir}.
    \end{equation*}

    Now consider the torsion subgroup of $S^1$. By definition, $z\in S^1$
    has finite order, say $n$, if and only if $z^n = 1$, i.e. if and
    only if $z\in G$. So we see that $G$ is actually the torsion
    subgroup of $S^1$.

    Since $\R/\Z$ is isomorphic to $S^1$, it follows that their
    torsion subgroups are also isomorphic. Hence $\Q/\Z\cong G$.
  \end{proof}
\end{enumerate}

\Exercise{15} Prove that a quotient of a divisible abelian group by
any proper subgroup is also divisible. Deduce that $\Q/\Z$ is
divisible.
\begin{proof}
  Let $G$ be a divisible abelian group and let $H<G$ be a proper
  subgroup. Let $g\in G$ be arbitrary and let $k\neq0$ be an
  integer. Since $G$ is divisible, there is an element $x\in G$ such
  that $x^k = g$. Then, by
  Exercise~\ref{exercise:quotient-group:coset-exponent}, we know that
  \begin{equation*}
    (xH)^k = x^kH = gH,
  \end{equation*}
  which shows that $G/H$ is also divisible.

  We have shown in
  Exercise~\ref{exercise:generated-subgroups:divisible-group} that
  $\Q$ is divisible. Since $\Z$ is a proper subgroup, we must have
  that $\Q/\Z$ is divisible as well.
\end{proof}

\Exercise{16}
\label{exercise:quotient-group:generators}
Let $G$ be a group, let $N$ be a normal subgroup of $G$ and let
$\overline{G} = G/N$. Prove that if $G = \gen{x,y}$ then
$\overline{G} = \gen{\bar x,\bar y}$. Prove more generally that if
$G = \gen{S}$ for any subset $S$ of $G$, then
$\overline{G} = \gen{\overline{S}}$.
\begin{proof}
  We will prove the general case, since the arguments are
  similar. Suppose $G = \gen{S}$. Then for any $g\in G$, we may write
  \begin{equation*}
    g = s_1s_2\cdots s_k,
    \quad\text{where $s_i\in S$ for $1\leq i\leq k$}.
  \end{equation*}
  Let $\overline{S}$ be cosets of the form $sN$, where $s\in S$. Then
  $gN$ may be written as
  \begin{align*}
    gN &= (s_1s_2\cdots s_k)N \\
       &= (s_1N)(s_2N)\cdots(s_kN).
  \end{align*}
  This shows that every coset in $\overline{G}$ can be written as a
  product of cosets in $\overline{S}$. Therefore
  $\overline{G} = \gen{\overline{S}}$.
\end{proof}

\Exercise{17} Let $G$ be the dihedral group of order $16$:
\begin{equation*}
  G = \gen{r,s\mid r^8 = s^2 = 1, rs = sr^{-1}}
\end{equation*}
and let $\overline{G} = G/\gen{r^4}$ be the quotient of $G$ by the
subgroup generated by $r^4$ (this subgroup is the center of $G$, hence
is normal).
\begin{enumerate}
\item Show that the order of $\overline{G}$ is $8$.
  \begin{solution}
    $\gen{r^4} = \{1, r^4\}$. So for each $g\in G$, we can write
    $g\gen{r^4} = \{g, gr^4\}$, and we see that the cosets in
    $\overline{G}$ partition $G$ into sets of two elements.
    $\ord{G} = 16$, so there must be $8$ distinct cosets in
    $\overline{G}$.
  \end{solution}
\item Exhibit each element of $\overline{G}$ in the form
  $\bar s^a\bar r^b$, for some integers $a$ and $b$.
  \begin{solution}
    The elements of $\overline{G}$ are
    \begin{align*}
      \bar1 &= \{1, r^4\}, & \bar{s} &= \{s, sr^4\}, \\
      \bar{r} &= \{r, r^5\}, & \bar{s}\bar{r} &= \{sr, sr^5\}, \\
      \bar{r}^2 &= \{r^2, r^6\}, & \bar{s}\bar{r}^2 &= \{sr^2, sr^6\}, \\
      \bar{r}^3 &= \{r^3, r^7\}, & \bar{s}\bar{r}^3 &= \{sr^3, sr^7\}.
                                                      \qedhere
    \end{align*}
  \end{solution}
\item Find the order of each of the elements of $\overline{G}$
  exhibited in (b).
  \begin{solution}
    The orders are
    \begin{align*}
      \ord{\bar1} &= 1, & \ord{\bar{s}} &= 2, \\
      \ord{\bar{r}} &= 4, & \ord{\overline{sr}} &= 2, \\
      \ord{\overline{r^2}} &= 2, & \ord{\overline{sr^2}} &= 2, \\
      \ord{\overline{r^3}} &= 4, & \ord{\overline{sr^3}} &= 2. \qedhere
    \end{align*}
  \end{solution}
\item Write each of the following elements of $\overline{G}$ in the
  form $\bar s^a\bar r^b$, for some integers $a$ and $b$ as in (b):
  $\overline{rs}$, $\overline{sr^{-2}s}$, $\overline{s^{-1}r^{-1}sr}$.
  \begin{solution}
    Since $rs = sr^{-1} = sr^7$, and since
    $\overline{sr^7} = \overline{sr^3}$, we have
    $\overline{rs} = \overline{sr^3}$. Likewise,
    \begin{equation*}
      \overline{sr^{-2}s} = \overline{s^2r^2} = \overline{r^2}
    \end{equation*}
    and
    \begin{equation*}
      \overline{s^{-1}r^{-1}sr}
      = \overline{s^{-1}sr^2}
      = \overline{r^2}. \qedhere
    \end{equation*}
  \end{solution}
\item Prove that $\overline{H} = \gen{\bar s, \bar r^2}$ is a normal
  subgroup of $\overline{G}$ and $\overline{H}$ is isomorphic to the
  Klein $4$-group. Describe the isomorphism type of the complete
  preimage of $\overline{H}$ in $G$.
  \begin{solution}
    By definition, $\overline{H}$ is a subgroup of $\overline{G}$
    since it is generated by members of $\overline{G}$. To show that
    $\overline{H}\trianglelefteq\overline{G}$, we need only check that
    the conjugates of the generators of $\overline{H}$ lie in
    $\overline{H}$.

    Take any element $\bar{g}\in\overline{G}$. If
    $\bar{g} = \overline{r^k}$ for some integer $k$ with $0\leq k\leq 3$,
    then
    \begin{equation*}
      \overline{gsg^{-1}}
      = \overline{r^ksr^{-k}}
      = \overline{sr^{-2k}}
      = \overline{s(r^2)^{-k}}\in\overline{H}
    \end{equation*}
    and
    \begin{equation*}
      \overline{gr^2g^{-1}}
      = \overline{r^kr^2r^{-k}}
      = \overline{r^2}\in\overline{H}.
    \end{equation*}
    The other possibility is that $\bar{g} = \overline{sr^k}$. Then
    \begin{equation*}
      \overline{gsg^{-1}}
      = \overline{sr^ks(sr^k)^{-1}}
      = \overline{sr^kssr^k}
      = \overline{sr^{2k}}
      = \overline{s(r^2)^k}\in\overline{H}
    \end{equation*}
    and
    \begin{equation*}
      \overline{gr^2g^{-1}}
      = \overline{sr^kr^2(sr^k)^{-1}}
      = \overline{sr^{k+2}sr^k}
      = \overline{r^{-2}}
      = \overline{(r^2)^{-1}}\in\overline{H}.
    \end{equation*}
    This shows that $\overline{H}$ is normal in $\overline{G}$.

    Note that
    $\overline{H} = \{1, \overline{r^2}, \bar{s}, \overline{sr^2}\}$
    has exactly four elements. It has already been shown that every
    group having four elements is isomorphic to either the Klein
    4-group $V_4$ or to the cyclic group $Z_4$ (see
    Exercise~\ref{exercise:classify:groups-4}). $Z_4$ has only one
    element of order $2$, but it is easy to check that every
    nonidentity element in $\overline{H}$ has order $2$, so
    $\overline{H}\cong V_4$.

    Lastly, the complete preimage of $\overline{H}$ is
    $\pi^{-1}(\overline{H})$, where $\pi\colon G\to\overline{G}$ is
    the natural projection of $G$ onto $\overline{G}$. From the cosets
    found earlier, we find that
    \begin{equation*}
      \pi^{-1}(\overline{H})
      = \{1, r^2, r^4, r^6, s, sr^2, sr^4, sr^6\}.
    \end{equation*}
    Call this preimage $A$. If we notice that
    \begin{equation*}
      (r^2)^4 = s^2 = 1
      \quad\text{and}\quad
      r^2s = sr^{-2} = s(r^2)^{-1},
    \end{equation*}
    we see that this subgroup $A$ behaves like $D_8$, and indeed, the
    mapping $\varphi\colon A\to D_8$ determined by $\varphi(r^2) = r$
    and $\varphi(s) = s$ extends to a bijective
    homomorphism. Therefore $A\cong D_8$.
  \end{solution}
\item Find the center of $\overline{G}$ and describe the isomorphism
  type of $\overline{G}/Z(\overline{G})$.
  \begin{solution}
    From Exercise~\ref{exercise:quotient-group:generators}, we know
    that $\overline{G} = \gen{\bar{r}, \bar{s}}$. Since
    $\overline{r^2}$ commutes with both of these generators,
    $\overline{r^2}\in Z(\overline{G})$. However, the elements
    $\bar{r}$ and $\overline{r^3}$ do not commute with $s$, and the
    elements $\overline{sr}$, $\overline{sr^2}$, and $\overline{sr^3}$
    do not commute with $r$. So
    \begin{equation*}
      Z(\overline{G}) = \{ \bar1, \overline{r^2} \}.
    \end{equation*}

    The elements of $\overline{G}/Z(\overline{G})$ are
    \begin{align*}
      \overline{\overline1}
      &= \{\bar1, \overline{r^2}\},
      & \overline{\overline{s}}
      &= \{\bar{s}, \overline{sr^2}\}, \\
      \overline{\overline{r}}
      &= \{\bar{r}, \overline{r^3}\},
      & \overline{\overline{sr}}
      &= \{\overline{sr}, \overline{sr^3}\}.
    \end{align*}
    Notice that each nonidentity element has order $2$. Therefore
    $\overline{G}/Z(\overline{G})$ is isomorphic to the Klein 4-group,
    $V_4$.
  \end{solution}
\end{enumerate}

\Exercise{18} Let $G$ be the quasidihedral group of order $16$:
\begin{equation*}
  G = \gen{\sigma, \tau\mid\sigma^8=\tau^2=1, \sigma\tau=\tau\sigma^3}
\end{equation*}
and let $\overline{G} = G/\gen{\sigma^4}$ be the quotient of $G$ by
the subgroup generated by $\sigma^4$ (this subgroup is the center of
$G$, hence is normal).
\begin{enumerate}
\item Show that the order of $\overline{G}$ is $8$.
  \begin{solution}
    Since $\gen{\sigma^4} = \{1,\sigma^4\}$, its cosets partition $G$
    into pairs of elements, so that there must be $8$ distinct cosets.
  \end{solution}

\item Exhibit each element of $\overline{G}$ in the form
  $\bar\tau^a\bar\sigma^b$, for some integers $a$ and $b$.
  \begin{solution}
    The elements of $\overline{G}$ are as follows.
    \begin{align*}
      \bar1 &= \{1, \sigma^4\},
      & \bar{\tau} &= \{\tau, \tau\sigma^4\}, \\
      \bar\sigma &= \{\sigma, \sigma^5\},
      & \overline{\tau\sigma} &= \{\tau\sigma, \tau\sigma^5\}, \\
      \overline{\sigma^2} &= \{\sigma^2, \sigma^6\},
      & \overline{\tau\sigma^2} &= \{\tau\sigma^2, \tau\sigma^6\}, \\
      \overline{\sigma^3} &= \{\sigma^3, \sigma^7\},
      & \overline{\tau\sigma^3} &= \{\tau\sigma^3, \tau\sigma^7\}. \qedhere
    \end{align*}
  \end{solution}

\item Find the order of each of the elements of $\overline{G}$
  exhibited in (b).
  \begin{solution}
    Computing the orders of the elements, we find
    \begin{align*}
      \ord{\bar1} &= 1,
      & \ord{\bar\tau} &= 2, \\
      \ord{\bar\sigma} &= 4,
      & \ord{\overline{\tau\sigma}} &= 2, \\
      \ord{\overline{\sigma^2}} &= 2,
      & \ord{\overline{\tau\sigma^2}} &= 2, \\
      \ord{\overline{\sigma^3}} &= 4,
      & \ord{\overline{\tau\sigma^3}} &= 2. \qedhere
    \end{align*}
  \end{solution}

\item Write each of the following elements of $\overline{G}$ in the
  form $\bar\tau^a\bar\sigma^b$, for some integers $a$ and $b$ as in
  (b): $\overline{\sigma\tau}$, $\overline{\tau\sigma^{-2}\tau}$,
  $\overline{\tau^{-1}\sigma^{-1}\tau\sigma}$.
  \begin{solution}
    We have
    \begin{align*}
      \overline{\sigma\tau} &= \overline{\tau\sigma^3}, \\
      \overline{\tau\sigma^{-2}\tau} &= \overline{\sigma^2}, \\
      \intertext{and}
      \overline{\tau^{-1}\sigma^{-1}\tau\sigma}
      &= \overline{\sigma^6} = \overline{\sigma^2}.
      \qedhere
    \end{align*}
  \end{solution}

\item Prove that $\overline{G}\cong D_8$.
  \begin{proof}
    In $\overline{G}$, we know $\ord{\bar\tau} = 2$ and
    $\ord{\bar\sigma} = 4$, and we also know that
    \begin{equation*}
      \overline{\sigma\tau} = \overline{\tau\sigma^3}
      = \overline{\tau\sigma^7} = \overline{\tau\sigma^{-1}}.
    \end{equation*}
    So $\overline{G}$ satisfies the same relations as those given in
    the presentation of $D_8$ and has the same number of elements as
    $D_8$. Therefore $\overline{G}\cong D_8$.
  \end{proof}
\end{enumerate}

\Exercise{19} Let $G$ be the modular group of order $16$:
\begin{equation*}
  G = \gen{u,v \mid u^2 = v^8 = 1, vu = uv^5}
\end{equation*}
and let $\overline{G} = G/\gen{v^4}$ be the quotient of $G$ by the
subgroup generated by $v^4$ (this subgroup is contained in the center
of $G$, hence is normal).
\begin{enumerate}
\item Show that the order of $\overline{G}$ is $8$.
  \begin{solution}
    As in the previous two exercises, the cosets of
    $\gen{v^4} = \{1, v^4\}$ partition $\overline{G}$ into pairs of
    elements, so that $\overline{G} = 8$.
  \end{solution}
\item Exhibit each element of $\overline{G}$ in the form
  $\bar{u}^a\bar{v}^b$, for some integers $a$ and $b$.
  \begin{solution}
    The elements are
    \begin{align*}
      \bar1 &= \{1, v^4\}, & \bar{u} &= \{u, uv^4\}, \\
      \bar{v} &= \{v, v^5\}, & \bar{u}\bar{v} &= \{uv, uv^5\}, \\
      \bar{v}^2 &= \{v^2, v^6\}, & \bar{u}\bar{v}^2 &= \{uv^2, uv^6\}, \\
      \bar{v}^3 &= \{v^3, v^7\}, & \bar{u}\bar{v}^3 &= \{uv^3, uv^7\}.
                                                      \qedhere
    \end{align*}
  \end{solution}
\item Find the order of each of the elements of $\overline{G}$
  exhibited in (b).
  \begin{solution}
    The orders of the elements are
    \begin{align*}
      \ord{\bar1} &= 1,
      & \ord{\bar{u}} &= 2, \\
      \ord{\bar{v}} &= 4,
      & \ord{\overline{uv}} &= 4, \\
      \ord{\overline{v^2}} &= 2,
      & \ord{\overline{uv^2}} &= 2, \\
      \ord{\overline{v^3}} &= 4,
      & \ord{\overline{uv^3}} &= 4. \qedhere
    \end{align*}
  \end{solution}
\item Write each of the following elements of $\overline{G}$ in the
  form $\bar{u}^a\bar{v}^b$, for some integers $a$ and $b$ as in (b):
  $\overline{vu}$, $\overline{uv^{-2}u}$, $\overline{u^{-1}v^{-1}uv}$.
  \begin{solution}
    We have $\overline{vu} = \overline{uv^5} = \overline{uv}$,
    \begin{equation*}
      \overline{uv^{-2}u}
      = \overline{uv^6u}
      = \overline{vuvu}
      = \overline{vuuv^5}
      = \overline{v^6}
      = \overline{v^2},
    \end{equation*}
    and
    \begin{equation*}
      \overline{u^{-1}v^{-1}uv}
      = \overline{uv^7uv}
      = \overline{uv^5(v^2uv)}
      = \overline{vuuv^{10}v}
      = \overline{v^{12}}
      = \bar1. \qedhere
    \end{equation*}
  \end{solution}
\item Prove that $\overline{G}$ is abelian and is isomorphic to
  $Z_2\times Z_4$.
  \begin{proof}
    First, note that since
    $\overline{uv} = \overline{uv^5} = \overline{vu}$, the generators
    of $\overline{G}$ commute. Therefore $\overline{G}$ is abelian.

    Let $x$ be a generator for $Z_2$ and $y$ a generator for
    $Z_4$. Define the map $\varphi\colon\overline{G}\to Z_2\times Z_4$
    by
    \begin{equation*}
      \varphi(\bar{u}^a\bar{v}^b) = (x^a,y^b),
      \quad a,b\in\Z.
    \end{equation*}
    It is not hard to see that $\varphi$ is well defined, since in
    both groups the exponents $a$ and $b$ can be reduced modulo $2$ or
    $4$, respectively, in the same way.

    Now, let $\bar{g} = \bar{u}^a\bar{v}^b$ and
    $\bar{h} = \bar{u}^c\bar{v}^d$ be two elements of
    $\overline{G}$. Then
    \begin{align*}
      \varphi(\bar{g}\bar{h})
      &= \varphi(\overline{u^av^bu^cv^d}) \\
      &= \varphi(\overline{u^{a+c}v^{b+d}}) \\
      &= (x^{a+c},y^{b+d}) \\
      &= (x^a,y^b)(x^c,y^d) \\
      &= \varphi(\bar{g})\varphi(\bar{h}),
    \end{align*}
    and we see that $\varphi$ is a homomorphism.

    $\varphi$ is also surjective since each element of $Z^2\times Z^4$
    can be obtained with the appropriate exponents on $\bar{u}$ and
    $\bar{v}$. Being a surjective map between two sets of the same
    size, $\varphi$ must be a bijection, and thus an
    isomorphism. Hence $\overline{G}\cong Z_2\times Z_4$.
  \end{proof}
\end{enumerate}

\Exercise{20} Let $G = \Z/24\Z$ and let
$\widetilde{G} = G/\gen{\overline{12}}$, where for each integer $a$ we
simplify notation by writing $\widetilde{\overline{a}}$ as
$\tilde{a}$.
\begin{enumerate}
\item Show that
  $\widetilde{G} = \{\tilde0, \tilde1, \dots, \widetilde{11}\}$.
  \begin{proof}
    Since $\gen{\overline{12}} = \{\bar0, \overline{12}\}$, each coset
    will consist of the pair
    \begin{equation*}
      \{\bar{n}, \overline{12 + n}\} = \tilde{n}
      \quad\text{for $0\leq n\leq 11$},
    \end{equation*}
    and each of these is distinct.
  \end{proof}
\item Find the order of each element of $\widetilde{G}$.
  \begin{solution}
    The orders are
    \begin{align*}
      \ord{\tilde0} &= 1,
      & \ord{\tilde6} &= 2, \\
      \ord{\tilde1} &= 12,
      & \ord{\tilde7} &= 12, \\
      \ord{\tilde2} &= 6,
      & \ord{\tilde8} &= 3, \\
      \ord{\tilde3} &= 4,
      & \ord{\tilde9} &= 4, \\
      \ord{\tilde4} &= 3,
      & \ord{\widetilde{10}} &= 6, \\
      \ord{\tilde5} &= 12,
      & \ord{\widetilde{11}} &= 12. \qedhere
    \end{align*}
  \end{solution}
\item Prove that $\widetilde{G}\cong\Z/12\Z$.
  \begin{proof}
    Define the function $\varphi\colon\widetilde{G}\to\Z/12\Z$ by
    \begin{equation*}
      \varphi(\tilde{n}) = \bar{n}.
    \end{equation*}
    This is clearly a bijection, and it is a homomorphism since
    \begin{equation*}
      \varphi(\widetilde{m} + \tilde{n})
      = \overline{m + n}
      = \overline{m} + \bar{n}
      = \varphi(\widetilde{m}) + \varphi(\tilde{n})
    \end{equation*}
    for any $\widetilde{m},\tilde{n}\in\widetilde{G}$. Therefore
    $\widetilde{G}\cong\Z/12\Z$.
  \end{proof}
\end{enumerate}

\Exercise{21} Let $G = Z_4\times Z_4$ be given in terms of the
following generators and relations:
\begin{equation*}
  G = \gen{x,y \mid x^4 = y^4 = 1, xy = yx}.
\end{equation*}
Let $\overline{G} = G/\gen{x^2y^2}$ (note that every subgroup of the
abelian group $G$ is normal).
\begin{enumerate}
\item Show that the order of $\overline{G}$ is $8$.
  \begin{proof}
    Since $x$ and $y$ commute, $(x^2y^2)^2 = 1$ and so
    $\gen{x^2y^2} = \{1, x^2y^2\}$. The cosets of $\gen{x^2y^2}$ in
    $G$ are as follows:
    \begin{align*}
      \bar1 &= \{1, x^2y^2\},
      & \bar{y} &= \{y, x^2y^3\}, \\
      \bar{x} &= \{x, x^3y^2\},
      & \overline{xy} &= \{xy, x^3y^3\}, \\
      \overline{x^2} &= \{x^2, y^2\},
      & \overline{x^2y} &= \{x^2y, y^3\}, \\
      \overline{x^3} &= \{x^3, xy^2\},
      & \overline{x^3y} &= \{x^3y, xy^3\}.
    \end{align*}
    Each of these is distinct, and all elements in $G$ are accounted
    for. Therefore $\ord{\overline{G}} = 8$.
  \end{proof}
\item Exhibit each element of $\overline{G}$ in the form
  $\bar{x}^a\bar{y}^b$, for some integers $a$ and $b$.
  \begin{solution}
    As given above, the elements are $\bar1$, $\bar{x}$,
    $\overline{x^2}$, $\overline{x^3}$, $\bar{y}$, $\overline{xy}$,
    $\overline{x^2y}$, and $\overline{x^3y}$.
  \end{solution}
\item Find the order of each of the elements of $\overline{G}$
  exhibited in (b).
  \begin{solution}
    The orders are
    \begin{align*}
      \ord{\bar1} &= 1,
      & \ord{\bar{y}} &= 4, \\
      \ord{\bar{x}} &= 4,
      & \ord{\overline{xy}} &= 2, \\
      \ord{\overline{x^2}} &= 2,
      & \ord{\overline{x^2y}} &= 4, \\
      \ord{\overline{x^3}} &= 4,
      & \ord{\overline{x^3y}} &= 2. \qedhere
    \end{align*}
  \end{solution}
\item Prove that $\overline{G} \cong Z_4\times Z_2$.
  \begin{proof}
    Let $a$ be a generator for $Z_4$ and $b$ a generator for
    $Z_2$. Also, in $\overline{G}$, let $\bar{u} = \bar{x}$ and
    $\bar{v} = \overline{xy}$, and note that every element in
    $\overline{G}$ can be written in the form $\bar{u}^m\bar{v}^n$ for
    integers $m$ and $n$. Define a function
    $\varphi\colon\overline{G}\to Z_4\times Z_2$ by
    \begin{equation*}
      \varphi(\overline{u^mv^n}) = (a^m, b^n).
    \end{equation*}

    First we show that $\varphi$ is well defined. Suppose
    $\overline{u^mv^n} = \overline{u^pv^q}$. Then
    $\overline{x^{m+n}y^n} = \overline{x^{p+q}y^q}$ and we must have
    $n\equiv q\pmod4$ and $m+n\equiv p+q\pmod4$, which together imply
    that $m\equiv p\pmod4$ and $n\equiv q\pmod2$. This means that in
    $Z_4\times Z_2$, the elements $(a^m,b^n)$ and $(a^p,b^q)$ are
    actually the same, so $\varphi$ is a well defined function.

    Next, $\varphi$ is easily seen to be surjective. And since both
    groups have the same order, it is also a bijection.

    Finally, $\varphi$ is a homomorphism since
    \begin{align*}
      \varphi((a^m,b^n)(a^p,b^q))
      &= \varphi((a^{m+p},b^{n+q})) \\
      &= \overline{x^{m+p}y^{n+q}} \\
      &= \overline{x^my^n}\cdot\overline{x^py^q} \\
      &= \varphi((a^m,b^n))\varphi((a^p,b^q)).
    \end{align*}
    Therefore $\overline{G}\cong Z_4\times Z_2$.
  \end{proof}
\end{enumerate}

\Exercise{22}
\begin{enumerate}
\item Prove that if $H$ and $K$ are normal subgroups of a group $G$
  then their intersection $H\cap K$ is also a normal subgroup of $G$.
  \begin{proof}
    Suppose $H$ and $K$ are normal subgroups of $G$. We already know
    that $H\cap K$ is a subgroup of $G$, so we need to show that it is
    normal. Choose any $g\in G$ and any $x\in H\cap K$. Since $x\in H$
    and $H\trianglelefteq G$, we know $gxg^{-1}\in H$. Likewise, since
    $x\in K$ and $K\trianglelefteq G$, we have $gxg^{-1}\in
    K$. Therefore $gxg^{-1}\in H\cap K$. This shows that
    $g(H\cap K)g^{-1}\subseteq H\cap K$, and this is true for all
    $g\in G$. By Theorem~6~(5) (which we will prove in
    Exercise~\ref{exercise:quotient-group:normal-criterion}), this is
    enough to show that $H\cap K\trianglelefteq G$.
  \end{proof}
\item Prove that the intersection of an arbitrary nonempty collection
  of normal subgroups of a group is a normal subgroup (do not assume
  the collection is countable).
  \begin{proof}
    Let $G$ be a group and let $I$ be a nonempty set of indices, not
    necessarily countable. Consider the collection of subgroups
    $\{N_\alpha \mid \alpha\in I\}$, where $N_\alpha\trianglelefteq G$
    for each $\alpha\in I$. Let
    \begin{equation*}
      N = \bigcap_{\alpha\in I}N_\alpha.
    \end{equation*}
    We know $N$ is a subgroup of $G$ by
    Exercise~\ref{exercise-intersection-of-subgroups-is-subgroup}.

    For any $g\in G$ and any $n\in N$, we must have $n\in N_\alpha$
    for each $\alpha$. And since $N_\alpha\trianglelefteq G$, we have
    $gng^{-1}\in N_\alpha$ for each $\alpha$. Therefore
    $gng^{-1}\in N$, which shows that $gNg^{-1}\subseteq N$ for each
    $g\in G$. As before, this is enough to complete the proof.
  \end{proof}
\end{enumerate}

\Exercise{23} Prove that the join of any nonempty collection of normal
subgroups of a group is a normal subgroup.
\begin{proof}
  Let $G$ be a group and $I$ a nonempty set of indices, and let
  $\{N_\alpha\mid\alpha\in I\}$ be a collection of subgroups of $G$,
  with $N_\alpha\trianglelefteq G$ for each $\alpha\in I$. Let $N$ be
  the join of all the subgroups in the collection:
  \begin{equation*}
    N = \gen{N_\alpha \mid \alpha\in I}.
  \end{equation*}

  Fix an element $g\in G$ and $n\in N$. Since $n$ belongs to the join
  of $\{N_\alpha\}$, we can write $n$ as the product of finitely many
  elements, each belonging to one of the sets in the collection:
  \begin{equation*}
    n = x_1x_2\cdots x_k,
    \quad\text{where for each $i$,
      $x_i\in N_{\alpha_i}$ for some $\alpha_i\in I$}.
  \end{equation*}
  Since $N_{\alpha_i}\trianglelefteq G$ for each $i$, we have
  $gx_ig^{-1}\in N_{\alpha_i}\subseteq N$. Now notice that we can
  write
  \begin{align*}
    gng^{-1}
    &= gx_1x_2x_3\cdots x_{k-1}x_kg^{-1} \\
    &= gx_1(g^{-1}g)x_2(g^{-1}g)x_3\cdots x_{k-1}(g^{-1}g)x_kg^{-1} \\
    &= (gx_1g^{-1})(gx_2g^{-1})\cdots(gx_kg^{-1}).
  \end{align*}
  Therefore $gng^{-1}$ can be written as a product of elements in $N$,
  and so must be in $N$ by closure. Since $g$ and $n$ were chosen
  arbitrarily, this shows that $gNg^{-1}\subseteq N$ for each $g\in G$
  so that $N\trianglelefteq G$.
\end{proof}

\Exercise{24} Prove that if $N\trianglelefteq G$ and $H$ is any
subgroup of $G$ then $N\cap H\trianglelefteq H$.
\begin{proof}
  We know that the intersection of two subgroups is a subgroup, so
  $N\cap H$ is a subgroup of $H$. Let $h\in H$ and $x\in N\cap
  H$. Then since $x\in N$ and $N\trianglelefteq G$, $hxh^{-1}\in
  N$. But $x\in H$ so $hxh^{-1}\in H$. Therefore
  $hxh^{-1}\in N\cap H$. Since this is true for any $h\in H$ and
  $x\in N\cap H$, we have $N\cap H\trianglelefteq H$.
\end{proof}

\Exercise{25}
\label{exercise:quotient-group:normal-criterion}
\begin{enumerate}
\item Prove that a subgroup $N$ of $G$ is normal if and only if
  $gNg^{-1}\subseteq N$ for {\em all} $g\in G$.
  \begin{proof}
    The left-to-right implication is immediate from the
    definition. For the other direction, suppose that $N$ is a
    subgroup of a group $G$, with
    \begin{equation}
      \label{equation:quotient-group:g-normalize-N}
      gNg^{-1}\subseteq N
      \quad\text{for each $g\in G$}.
    \end{equation}
    Fix an element $g_0\in G$ and let $x\in N$. Consider the element
    $y = g_0^{-1}xg_0$. Then $y\in N$ by
    \eqref{equation:quotient-group:g-normalize-N} (simply take
    $g = g_0^{-1}$). Therefore
    \begin{equation*}
      g_0yg_0^{-1} = g_0(g_0^{-1}xg_0)g_0^{-1} = x.
    \end{equation*}
    This shows that $x\in g_0Ng_0^{-1}$. Since $g_0\in G$ was
    arbitrary, $x\in gNg^{-1}$ for each $g$ so that
    $N\subseteq gNg^{-1}$. Together with
    \eqref{equation:quotient-group:g-normalize-N}, we have
    $gNg^{-1} = N$ so that $N\trianglelefteq G$.
  \end{proof}
\item Let $G = GL_2(\Q)$, let $N$ be the subgroup of upper triangular
  matrices with integer entries and $1$'s on the diagonal, and let $g$
  be the diagonal matrix with entries $2,1$. Show that
  $gNg^{-1}\subseteq N$ but $g$ does {\em not} normalize $N$.
  \begin{proof}
    We have
    \begin{equation*}
      N = \left\{
        \begin{pmatrix}
          1 & b \\
          0 & 1
        \end{pmatrix}
        \;\middle|\;
        b\in\Z
      \right\}
      \quad\text{and}\quad
      g =
      \begin{pmatrix}
        2 & 0 \\
        0 & 1
      \end{pmatrix}.
    \end{equation*}
    For any $b\in\Z$,
    \begin{equation*}
      g\begin{pmatrix} 1 & b \\ 0 & 1 \end{pmatrix}g^{-1} =
      \begin{pmatrix}
        2 & 0 \\
        0 & 1
      \end{pmatrix}
      \begin{pmatrix}
        1 & b \\
        0 & 1
      \end{pmatrix}
      \begin{pmatrix}
        \frac12 & 0 \\
        0 & 1
      \end{pmatrix} \\
      =
      \begin{pmatrix}
        1 & 2b \\
        0 & 1
      \end{pmatrix}\in N,
    \end{equation*}
    and we see that $gNg^{-1}\subseteq N$.

    But notice that elements in $gNg^{-1}$ must have an even integer
    in the upper-right entry. This will always be the case, regardless
    of which $n\in N$ we use. Now consider the element
    \begin{equation*}
      x =
      \begin{pmatrix}
        1 & 1 \\
        0 & 1
      \end{pmatrix}\in N.
    \end{equation*}
    Then there is no matrix $n$ in $N$ with the property that
    $gng^{-1} = x$, because as we have observed, the upper-right entry
    of $gng^{-1}$ is always even. Consequently, $g$ does not normalize
    $N$.
  \end{proof}
\end{enumerate}

\Exercise{26} Let $a,b\in G$.
\begin{enumerate}
\item Prove that the conjugate of the product of $a$ and $b$ is the
  product of the conjugate of $a$ and the conjugate of $b$. Prove that
  the order of $a$ and the order of any conjugate of $a$ are the same.
  \begin{proof}
    For any $g\in G$, we have
    \begin{equation*}
      g(ab)g^{-1}
      = ga(g^{-1}g)bg^{-1}
      = (gag^{-1})(gbg^{-1}),
    \end{equation*}
    so the conjugate of a product is the product of the
    conjugates. Note that a simple induction argument will allow us to
    extend this result to products of any finite number of elements of
    $G$.

    Next, suppose $\ord{a} = n < \infty$. From the above
    result, we have
    \begin{equation*}
      (gag^{-1})^n = ga^ng^{-1} = gg^{-1} = 1,
    \end{equation*}
    so $\ord{gag^{-1}}\leq n$. On the other hand, if
    $(gag^{-1})^k = 1$, then $ga^kg^{-1} = 1$, and multiplying on the
    left by $g^{-1}$ and on the right by $g$ gives
    $a^k = g^{-1}g = 1$, so that $\ord{gag^{-1}}\geq n$. Therefore
    \begin{equation*}
      \ord{gag^{-1}} = n = \ord{a},
      \quad\text{for all $g\in G$}.
    \end{equation*}

    Lastly, if $\ord{a} = \infty$, then there is no positive integer
    $n$ such that $a^n = 1$. But if $(gag^{-1})^n = 1$ for some
    positive $n$, then the same argument as before shows that
    $a^n = 1$, which is a contradiction. Therefore
    $\ord{gag^{-1}} = \infty$ in this case.
  \end{proof}
\item Prove that the conjugate of $a^{-1}$ is the inverse of the
  conjugate of $a$.
  \begin{proof}
    For any $g\in G$,
    \begin{equation*}
      (ga^{-1}g^{-1})(gag^{-1})
      = ga^{-1}(g^{-1}g)ag^{-1}
      = ga^{-1}ag^{-1}
      = gg^{-1}
      = 1,
    \end{equation*}
    so $ga^{-1}g^{-1} = (gag^{-1})^{-1}$.
  \end{proof}
\item Let $N = \gen{S}$ for some subset $S$ of $G$. Prove that
  $N\trianglelefteq G$ if $gSg^{-1}\subseteq N$ for all $g\in G$.
  \begin{proof}
    First note that if $S$ is the empty set, then $N$ is the trivial
    subgroup and is therefore normal in $G$. So assume that $S$ is
    nonempty.

    Now suppose $gSg^{-1}\subseteq N$ for all $g\in G$, and pick any
    $x\in N$. Since $N = \gen{S}$, we may write
    \begin{equation*}
      x = s_1s_2\cdots s_k,
      \quad\text{where $s_i\in S$ for each $i = 1,2,\dots,k$}.
    \end{equation*}
    Since we have already proven above that the conjugate of a product
    is the product of the conjugates, we have for all $g\in G$ that
    \begin{equation*}
      gxg^{-1} = (gs_1g^{-1})(gs_2g^{-1})\cdots(gs_kg^{-1})\in N.
    \end{equation*}
    Therefore $gNg^{-1}\subseteq N$ for all $g\in G$ and we can
    conclude that $N\trianglelefteq G$.
  \end{proof}
\item Deduce that if $N$ is the cyclic group $\gen{x}$, then $N$ is
  normal in $G$ if and only if for each $g\in G$, $gxg^{-1} = x^k$ for
  some $k\in\Z$.
  \begin{proof}
    Since $a\in N$ if and only if $a = x^k$ for some integer $k$, the
    left-to-right implication is immediate from the definition of a
    normal subgroup, and the other direction is just a special case of
    the previous result, with $S = \{x\}$.
  \end{proof}
\item Let $n$ be a positive integer. Prove that the subgroup $N$ of
  $G$ generated by all the elements of $G$ of order $n$ is a normal
  subgroup of $G$.
  \begin{proof}
    Let $S = \{g\in G\mid\ord{g} = n\}$ and let $N = \gen{S}$. If $S$
    is empty then $N$ is the trivial subgroup and is normal in $G$, so
    assume $S$ is nonempty. Then for any $s\in S$ and $g\in G$, we
    have
    \begin{equation*}
      \ord{gsg^{-1}} = \ord{s} = n,
    \end{equation*}
    so $gsg^{-1}\in S\subseteq{N}$. Then $gSg^{-1}\subseteq N$ for
    each $g\in G$, and we can apply our earlier result to conclude
    that $N\trianglelefteq G$.
  \end{proof}
\end{enumerate}

\Exercise{27}
\label{exercise:quotient-group:normalizer-finite-criterion}
Let $N$ be a {\em finite} subgroup of a group $G$. Show
that $gNg^{-1}\subseteq N$ if and only if $gNg^{-1} = N$. Deduce that
$N_G(N) = \{g\in G\mid gNg^{-1}\subseteq N\}$.
\begin{proof}
  Fix an element $g\in G$. If $gNg^{-1} = N$ then certainly
  $gNg^{-1}\subseteq N$.

  Conversely, suppose that $gNg^{-1}\subseteq N$. Define the function
  $\varphi\colon N\to gNg^{-1}$ by
  \begin{equation*}
    \varphi(x) = gxg^{-1}.
  \end{equation*}
  We will show that $\varphi$ is a bijection. First, if
  $gag^{-1} = gbg^{-1}$ for $a,b\in N$, then cancellation shows that
  $a = b$. Therefore $\varphi$ is injective. And if $y\in gNg^{-1}$,
  then by definition there is some $x\in N$ with
  $y = gxg^{-1} = \varphi(x)$, so $\varphi$ is surjective. We conclude
  that $\varphi$ is a bijection, and therefore
  $\ord{gNg^{-1}} = \ord{N}$. But $gNg^{-1}\subseteq N$ and $N$ is
  finite, so we must have equality:
  \begin{equation*}
    gNg^{-1} = N.
  \end{equation*}

  Finally, we know by definition that
  $N_G(N) = \{g\in G\mid gNg^{-1} = N\}$. As was just proved,
  $gNg^{-1} = N$ if and only if $gNg^{-1}\subseteq N$ (since $N$ is
  finite), so the normalizer can also be written as
  \begin{equation*}
    N_G(N) = \{g\in G\mid gNg^{-1}\subseteq N\}. \qedhere
  \end{equation*}
\end{proof}

\Exercise{28}
\label{exercise:quotient-group:normalizer-iff-generator1}
Let $N$ be a {\em finite} subgroup of a group $G$ and
assume $N = \gen{S}$ for some subset $S$ of $G$. Prove that an element
$g\in G$ normalizes $N$ if and only if $gSg^{-1}\subseteq N$.
\begin{proof}
  Fix an element $g\in G$. First, if $gNg^{-1} = N$ then it must be
  true that $gSg^{-1}\subseteq N$, since $gSg^{-1}$ is a subset of
  $gNg^{-1}$.

  Conversely, suppose $gSg^{-1}\subseteq N$. If $S$ is empty then $N$
  is trivial and must be normal, so suppose $S$ is nonempty. Choose
  any $x\in N$. Since $S$ generates $N$, we have
  \begin{equation*}
    x = s_1s_2\cdots s_k,
    \quad\text{where $s_i\in S$ for each $i=1,2,\dots,k$}.
  \end{equation*}
  As the conjugate of a product is the product of the conjugates, we
  know that $gxg^{-1} = gs_1g^{-1}\cdots gs_kg^{-1}$. Since
  $gs_ig^{-1}\in gSg^{-1}\subseteq N$, we have $gxg^{-1}\in N$. And
  our choice of $x\in N$ was arbitrary, so $gNg^{-1}\subseteq
  N$. Since $N$ is finite, we may use the result from
  Exercise~\ref{exercise:quotient-group:normalizer-finite-criterion},
  to conclude that $g$ normalizes $N$.
\end{proof}

\Exercise{29} Let $N$ be a {\em finite} subgroup of $G$ and suppose
$G = \gen{T}$ and $N = \gen{S}$ for some subsets $S$ and $T$ of
$G$. Prove that $N$ is normal in $G$ if and only if
$tSt^{-1}\subseteq N$ for all $t\in T$.
\begin{proof}
  If $N$ is normal in $G$ then $gNg^{-1} = N$ for all $g\in G$, so
  clearly $tSt^{-1}\subseteq N$ for any $t\in T$.

  For the other direction, suppose that $tSt^{-1}\subseteq N$ for all
  $t\in T$. If $S$ or $T$ is the empty set, then the result is
  obvious, so suppose $S$ and $T$ are nonempty. Choose any $g\in
  G$. We can write
  \begin{equation*}
    g = t_1t_2\cdots t_k
    \quad\text{with $t_i\in T$ for each $i = 1,2,\dots,k$}.
  \end{equation*}

  We will use induction on $k$ to prove that $gSg^{-1}\subseteq N$. If
  $k = 1$, then $gSg^{-1} = t_1St_1^{-1} \subseteq N$, so the base
  case is satisfied. Now assume that $gSg^{-1}\subseteq N$ whenever
  $g$ can be written as the product of $k$ elements from $t$, and
  consider
  \begin{equation*}
    g = t_1t_2\cdots t_kt_{k+1},
    \quad\text{$t_i\in T$ for each $i$}.
  \end{equation*}
  Set $h = t_1t_2\cdots t_k$, so that $g = ht_{k+1}$. By the induction
  assumption, $hSh^{-1}\subseteq N$. So for any $s\in S$, we have
  \begin{equation*}
    gsg^{-1}
    = ht_{k+1}s(ht_{k+1})^{-1}
    = ht_{k+1}st_{k+1}^{-1}h^{-1}
    = hxh^{-1},
  \end{equation*}
  where $x = t_{k+1}st_{k+1}^{-1}\in N$. So $x$ can be written as
  \begin{equation*}
    x = s_1s_2\cdots s_\ell,
    \quad\text{$s_i\in S$ for $i = 1, 2, \dots, \ell$}
  \end{equation*}
  and
  \begin{equation*}
    hxh^{-1} = (hs_1h^{-1})(hs_2h^{-1})\cdots(hs_\ell h^{-1}) \in N.
  \end{equation*}
  So $gsg^{-1}\in N$, which gives $gSg^{-1}\subseteq N$. By induction,
  this statement is true for any $g\in G$. And since $N$ is finite,
  our result from
  Exercise~\ref{exercise:quotient-group:normalizer-iff-generator1}
  finishes the proof.
\end{proof}

\Exercise{30} Let $N\leq G$ and let $g\in G$. Prove that $gN = Ng$ if
and only if $g\in N_G(N)$.
\begin{proof}
  Suppose $gN = Ng$. Then for any $x\in N$, there is a $y\in N$ such
  that $gx = yg$. Multiplying on the right by $g^{-1}$ gives
  $gxg^{-1} = y\in N$. This is true for any $x\in N$, so
  $gNg^{-1}\subseteq N$. On the other hand, if $x\in N$, then there is
  a $y\in N$ such that $xg = gy$, and multiplying on the right by
  $g^{-1}$ gives $x = gyg^{-1}\in gNg^{-1}$. So $N\subseteq gNg^{-1}$
  and we conclude that the two sets are equal. Therefore
  $g\in N_G(N)$.

  Conversely, suppose $g$ normalizes $N$. Let $x\in N$ be
  arbitrary. Then we have $x\in gNg^{-1}$ so that $x = gyg^{-1}$ for
  some $y\in N$. Multiplying on the right by $g$ gives $xg =
  gy$. Therefore $xg\in gN$ for all $x\in N$, so $Ng\subseteq gN$. By
  a symmetric argument, we also have $gN\subseteq Ng$. Therefore the
  two sets are equal: $gN = Ng$.
\end{proof}

\Exercise{31} Prove that if $H\leq G$ and $N$ is a normal subgroup of
$H$ then $H\leq N_G(N)$. Deduce that $N_G(N)$ is the largest subgroup
of $G$ in which $N$ is normal (i.e., is the join of all subgroups $H$
for which $N\trianglelefteq H$).
\begin{proof}
  We already know that $N_G(N)$ is a subgroup, so we only need to
  establish that $H\subseteq N_G(N)$. But this is easy: Let $h\in
  H$. Since $N\trianglelefteq H$ we have $hNh^{-1} = N$. Therefore
  $h\in N_G(N)$. This shows that $H\leq N_G(N)$.

  Clearly $N\trianglelefteq N_G(N)$, and we have shown that every
  subgroup of $G$ in which $N$ is normal must be a subgroup of
  $N_G(N)$. This implies that $N_G(N)$ is the largest subgroup of $G$
  in which $N$ is normal.
\end{proof}
