\chapter{Preface}

This is an unofficial solution guide to the book {\em Abstract
  Algebra}, Third Edition, by David S. Dummit and Richard M. Foote. It
is intended for students who are studying algebra with Dummit and
Foote's text. I encourage students who use this guide to first attempt
each exercise for themselves before looking up the solution, as doing
exercises is an essential part of learning mathematics.

In writing this guide, I have avoided using techniques or results
before the point at which they are introduced in the text. My
solutions should therefore be accessible to someone who is reading
through Dummit and Foote for the first time.

Given the large number of exercises, errors are unavoidable in a work
such as this. I have done my best to proofread each solution, but
mistakes will make it through nonetheless. If you find one, please
feel free to tell me about it in an email:
\href{mailto:gkikola@gmail.com}{gkikola@gmail.com}. I appreciate any
corrections or feedback.

Please know that this guide is currently unfinished. I am slowly
working on adding the remaining chapters, but this will be done at my
own pace. If you need a solution to an exercise that I have not
included, try typing the problem statement into a web search engine
such as \href{https://www.google.com}{Google}; it is likely that
someone else has already posted a solution.

This guide is licensed under the Creative Commons
Attribution-ShareAlike 4.0 International License. To view a copy of
this license, visit
\begin{center}
  \href{http://creativecommons.org/licenses/by-sa/4.0/}
  {http://creativecommons.org/licenses/by-sa/4.0/}
\end{center}

I am deeply grateful to the authors, David S. Dummit and Richard
M. Foote, for producing a wonderfully comprehensive and thoroughly
well-written work. I also owe particular thanks to those readers who
have taken the time to contact me in order to point out mistakes or to
give general feedback.

\begin{flushright}
  Greg Kikola\\
  \href{https://www.gregkikola.com/}{www.gregkikola.com}\\
  \href{mailto:gkikola@gmail.com}{gkikola@gmail.com}
\end{flushright}
